---
title: "Data Science Final Project"
knit: (function(input_file, encoding) {
    out_dir <- 'docs';
    rmarkdown::render(input_file,
      encoding=encoding,
      output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
author: "Vibha Vijay, Anish Mandalika, Salem Keleta"
date: "4/28/2022"
output:
  prettydoc::html_pretty:
    theme: cayman
    toc: true
    highlight: github
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### loading in the data and packages needed
```{r}
library(tidyverse)
library(caret)
library(rpart.plot)
library(randomForest)

fires = read.csv("Historical_Fire_Data.csv")
```

### casting the empties as NAs
```{r}
str(fires)
fires[fires ==" "] <- NA
```

### getting rid of unneeded columns
```{r}
fires = select(fires, -c(1,2,3,4,16,17,18,19,20))
```

### turning necessary columns into factors
```{r}
str(fires)
fires[,c(2,3,4,5,7,11,12)] <- lapply(fires[,c(2,3,4,5,7,11,12)], as.factor)

```

### Converting datetime string into season and time of day
```{r}
# Get a POSIXct datetime object from the string in the dataset
fires$AlarmDateTimeAmericaNY <- as.POSIXct(fires$AlarmDateTimeAmericaNY , format="%Y/%m/%d %H:%M:%S+00", tz="EST")

# Extract the date from the POSIXct object, convert it to a factor and collapse the factor into seasons
fires$season <- format(as.Date(fires$AlarmDateTimeAmericaNY), "%m")
fires$season <- as.factor(fires$season)
fires$season <- fct_collapse(fires$season,
                             Winter = c("01", "02", "03"),
                             Spring = c("04", "05", "06"),
                             Summer = c("07", "08", "09"),
                             Fall = c("10", "11", "12"))

# Extract the time from the POSIXct object, convert it to a factor and collapse the factor into seasons
fires$time <- format(fires$AlarmDateTimeAmericaNY, "%H")
fires$time <- as.factor(fires$time)
fires$time <- fct_collapse(fires$time,
                           Morning = c("06", "07", "08", "09", "10", "11"),
                           Afternoon = c("12", "13", "14", "15", "16", "17"),
                           Evening = c("18", "19", "20", "21", "22", "23"),
                           Night = c("00", "01", "02", "03", "04", "05"))

# Drop the original datetime variable since it's no longer neccesary
fires <- fires[, -1]
```

### Missing data
```{r}
mice::md.pattern(fires, rotate.names=TRUE)

# Fill in all NAs in factors with the string "Unknown"
fires[,c(1,2,3,4,6,10)] <- lapply(fires[,c(1,2,3,4,6,10)], addNA)
```

### Collapsing more factor levels
```{r}
fires$FireAreaOfOrigin <- fct_collapse(fires$FireAreaOfOrigin,
                                       `Small enclosed area` = c("Attic: vacant, crawl space above top story, cupola",
                                                               "Bathroom, checkroom, lavatory, locker room",
                                                               "Bedroom - < 5 persons; included are jail or prison",
                                                               "Ceiling & floor assembly, crawl space b/t stories",
                                                               "Chimney (conversion only)",
                                                               "Chute/container - trash, rubbish, waste",
                                                               "Closet",
                                                               "Common room, den, family room, living room, lounge",
                                                               "Conduit, pipe, utility, or ventilation shaft",
                                                               "Cooking area, kitchen",
                                                               "Duct: hvac, cable, exhaust, heating, or AC",
                                                               "Egress/exit, other",
                                                               "Entranceway, lobby",
                                                               "Heating room or area, water heater area",
                                                               "Interior stairway or ramp",
                                                               "Laboratory",
                                                               "Laundry area, wash house (laundry)",
                                                               "Light shaft",
                                                               "Machinery room or area; elevator machinery room",
                                                               "Office",
                                                               "Records storage room, storage vault",
                                                               "Sales area, showroom (exclude display window)",
                                                               "Sales area, showroom (excludes display window)",
                                                               "Storage room, area, tank, or bin",
                                                               "Storage: supplies or tools; dead storage",
                                                               "Structural area, other",
                                                               "Substructure area or space, crawl space",
                                                               "Surgery area - major operations, operating room",
                                                               "Switchgear area, transformer vault",
                                                               "Vacant structural area",
                                                               "Wall assembly",
                                                               "Wall assembly, concealed wall space"),
                                       `Large enclosed area` = c("Arena, assembly area w/ fixed seats - 100+ persons",
                                                                 "Art gallery, exhibit hall, library",
                                                                 "Assembly area - less than 100 persons",
                                                                 "Assembly area without fixed seats - 100+ persons",
                                                                 "Bedroom - 5+ persons; including barrack/dormitory",
                                                                 "Corridor, mall",
                                                                 "Dining room, cafeteria, bar area, beverage service",
                                                                 "Equipment or service area, other",
                                                                 "Function area, other",
                                                                 "Function areas, other",
                                                                 "Hallway corridor, mall",
                                                                 "Maintenance shop or area, paint shop or area",
                                                                 "Storage area, other",
                                                                 "Service facilities, other",
                                                                 "Vehicle storage area; garage, carport"),
                                       `Exterior in/near a building` = c("Awning",
                                                                         "Construction/renovation area",
                                                                         "Courtyard, patio, porch, terrace",
                                                                         "Courtyard, patio, terrace",
                                                                         "Exterior balcony, unenclosed porch",
                                                                         "Exterior stairway, ramp, or fire escape",
                                                                         "Exterior, exposed surface",
                                                                         "Incinerator area",
                                                                         "Roof surface: exterior",
                                                                         "Shipping/receiving area; loading area, dock or bay",
                                                                         "Wall surface: exterior"),
                                       `Open exterior area` = c("Highway, parking lot, street: on or near",
                                                                "Open area - outside; included are farmland, field",
                                                                "Open area, outside; included are farmland, field",
                                                                "Outside area, other",
                                                                "Railroad right of way: on or near",
                                                                "Wildland, woods"),
                                       `Vehicle` = c("Cargo/trunk area - all vehicles",
                                                     "Engine area, running gear, wheel area",
                                                     "Fuel tank, fuel line",
                                                     "Operator/passenger area of transportation equip.",
                                                     "Separate operator/control area of transportation",
                                                     "Vehicle area, other"),
                                       `Other` = c("Multiple areas",
                                                   "Other",
                                                   "Other area of fire origin",
                                                   "Undetermined",
                                                   NA))
fires$FireType <- fct_collapse(fires$FireType,
                               `Natural Material Fire` = c("Brush or brush-and-grass mixture fire",
                                                      "Cultivated vegetation, crop fire",
                                                      "Forest, woods or wildland fire",
                                                      "Grass fire",
                                                      "Natural vegetation fire",
                                                      "Natural vegetation fire, other"),
                               `Rubbish Fire` = c("Dumpster or other outside trash receptacle fire",
                                                  "Outside rubbish fire",
                                                  "Outside rubbish fire, other",
                                                  "Outside rubbish, trash or waste fire",
                                                  "Trash or rubbish fire, contained"),
                               `Structure Fire` = c("Building fire",
                                                   "Chimney or flue fire, confined to chimney or flue",
                                                   "Cooking fire, confined to container",
                                                   "Fire in mobile property used as a fixed structure",
                                                   "Fires in structure other than in a building",
                                                   "Fuel burner/boiler malfunction, fire confined",
                                                   "Structure Fire"),
                               `Vehicle Fire` = c("Fire in mobile home used as fixed residence",
                                                  "Mobile property (vehicle) fire",
                                                  "Mobile property (vehicle) fire, other",
                                                  "Passenger vehicle fire",
                                                  "Road freight or transport vehicle fire"),
                               `Other` = c("Fire, other",
                                           "Outside equipment fire",
                                           "Special outside fire",
                                           "Special outside fire, other",
                                           NA))
fires$HeatSource <- fct_collapse(fires$HeatSource,
                                 `Spark/Ember/Arc` = c("Arcing",
                                                       "Electrical arcing",
                                                       "Flying brand, ember, spark",
                                                       "Heat, spark from friction",
                                                       "Hot ember or ash",
                                                       "Lightning",
                                                       "Spark, ember, or flame from operating equipment"),
                                 `Existing fire/heat source` = c("Backfire from internal combustion engine",
                                                                 "Candle",
                                                                 "Flame/torch used for lighting",
                                                                 "Heat from direct flame, convection currents",
                                                                 "Heat from other open flame or smoking materials",
                                                                 "Heat from other open flame or smoking materials, other",
                                                                 "Hot or smoldering object, other",
                                                                 "Molten, hot material",
                                                                 "Radiated heat from another fire",
                                                                 "Radiated or conducted heat from operating equipment",
                                                                 "Radiated, conducted heat from operating equipment"),
                                 `Reactive/Inflammable/Explosive material` = c("Chemical reaction",
                                                                               "Chemical, natural heat source, other",
                                                                               "Cigarette",
                                                                               "Cigarette lighter",
                                                                               "Explosive, fireworks, other",
                                                                               "Fireworks",
                                                                               "Incendiary device",
                                                                               "Lighter: cigarette, cigar",
                                                                               "Match",
                                                                               "Pipe or cigar",
                                                                               "Spontaneous combustion, chemical reaction"),
                                 `Other` = c("Heat from powered equipment, other",
                                             "Heat from undetermined smoking material",
                                             "Heat source: other",
                                             "Multiple heat sources including multiple ignition",
                                             "Undetermined",
                                             NA))
fires$PropertyUse <- fct_collapse(fires$PropertyUse,  
                           Commercial_Building = c("Bank",
                                                   "Bar or nightclub",
                                                   "Business office",
                                                   "Communications center",
                                                   "Convenience store",
                                                    "Day care, in commercial property",
                                                    "Department or discount store",
                                                    "Doctor, dentist or oral surgeon's office",
                                                    "Eating, drinking places","Eating, drinking places, other",
                                                    "Food and beverage sales, grocery store",
                                                    "General retail, other","Hotel/motel, commercial","Household goods, sales, repairs","Laundry, dry cleaning","Live performance theater","Manufacturing, processing","Mercantile, business, other","Motor vehicle or boat sales, services, repair","Personal service, including barber & beauty shops","Professional supplies, services","Recreational, hobby, home repair sales, pet store","Restaurant or cafeteria","Service station, gas station","Specialty shop","Textile, wearing apparel sales"), 
                           Public_Facility = c("Adult education center, college classroom ","Assembly, other","Athletic/health club","Courthouse","Dump, sanitary landfill","Educational, other","Electrical distribution","Elementary school, including kindergarten","Fixed use recreation places, other","Fixed-use recreation places, other","Health care, detention, & correction, other","High school/junior high school/middle school","Playground","Preschool","Public or government, other","Reformatory, juvenile detention center","Schools, non-adult","Stadium, arena","Steam or heat generating plant","Storage, other","Water utility"),
                           Residence = c("1 or 2 family dwelling",
                                             "24-hour care Nursing homes, 4 or more persons","Barracks, dormitory", "Boarding/rooming house, residential hotels","Day care, in residence, licensed","Dormitory type residence, other","Dormitory-type residence, other","Multifamily dwelling","Parking garage, (detached residential garage)","Parking garage, general vehicle","Residential board and care","Residential or self storage units","Residential street, road or residential driveway","Residential, other","Sorority house, fraternity house"),
                           Outdoor = c("Beach",
                                             "Bridge, trestle",
                                             "Construction site","Forest, timberland, woodland","Graded and cared-for plots of land","Highway or divided highway","Industrial plant yard - area","Lake, river, stream","Open land or field","Outbuilding or shed","Outbuilding, protective shelter","Outside material storage area","Outside or special property, other","Pipeline, power line or other utility right of way","Railroad right of way","Railroad right-of-way","Railroad yard","Street or road in commercial area","Street, other","Tunnel","Undetermined","Vacant lot","Vehicle parking area","Vehicle storage, other"),
                           Private_Facility = c("Alcohol or substance abuse recovery center","Auditorium or concert hall","Ballroom, gymnasium","Church, mosque, synagogue, temple, chapel","Clinics, Doctors offices, hemodialysis centers","Hospital - medical or psychiatric","Laboratory or science lababoratory","Laboratory or science laboratory","None","Office: veterinary or research","Property Use, other","Refrigerated storage","Studio/theater, other","Swimming facility: indoor or outdoor","Utility or Distribution system, other","Variable use amusement, recreation places","Warehouse"),
                           Other = c(NA))
fires$CauseOfIgnition <- fct_collapse(fires$CauseOfIgnition,
                           Potential_Arson = c("Intentional"), 
                           Other_Cause = c("Albemarle County in charge of investigation","Cause under investigation","Cause undetermined after investigation","Cause, other","Cause, other (System generated code only, not used for data entry)","Other cause","Undetermined",NA),
                           Natural_Cause = c("Act of nature","Debris, vegetation burn","Natural source"),
                           Human_Accident = c("Equipment","Failure of equipment or heat source","Incendiary","Misuse of fire ","Open/outdoor fire","Smoking","Unintentional"))
fires$ItemFirstIgnited <- fct_collapse(fires$ItemFirstIgnited,  
                           Organic_Material = c("Agricultural crop, including fruits and vegetable","Animal, living or dead","Baled goods or material","Cabinetry (including built-in)","Chips, including wood chips","Cooking materials, including edible materials","Flammable liquid/gas - in/from engine or burner","Flammable liquid/gas - in/from final container","Flammable liquid/gas - uncontained","Flammable liquid/gas in container or pipe","Heavy vegetation - not crop, including trees","Light vegetation - not crop, including grass","Non-upholstered chair, bench","Organic materials, other","Pallet, skid (empty)"), 
                           Fiber = c("Awning, canopy","Bedding; blanket, sheet, comforter","Curtains, blinds, drapery, tapestry","Dust, fiber, lint, including sawdust and excelsior","Floor covering or rug/carpet/mat","Floor covering or rug/carpet/mat, surface","Goods not made up, including fabrics & yard goods","Goods not made up, including fabrics and yard goods","Linen; other than bedding","Magazine, newspaper, writing paper","Mattress, pillow","Rolled, wound material (paper, fabric)","Soft goods, wearing apparel, other","Tarpaulin, tent","Thermal, acoustical insulation within wall, partition or floor/ceiling space","Upholstered sofa, chair, vehicle seats","Wearing apparel not on a person","Wearing apparel on a person"),
                           Metal = c("Adornment, recreational material, signs, other","Appliance housing or casing","Conveyor belt, drive belt, V-belt","Electrical wire, cable insulation","Household utensils","Pipe, duct, conduit or hose","Railroad ties","Structural component or finish, other","Structural member or framing","Transformer, including transformer fluids"),
                           Other_Manmade_Material = c("Adhesive","Atomized liquid, vaporized liquid, aerosol.","Book","Box, carton, bag, basket, barrel","Decoration","Exterior roof covering or finish","Exterior sidewall covering, surface, finish","Exterior trim, including doors","Exterior wall covering or finish","Film, residue, including paint & resin","Filter, including evaporative cooler pads","Furniture, utensils, other","General materials, other","Insulation within structural area","Interior ceiling cover or finish","Interior ceiling covering or finish","Interior wall covering excluding drapes, etc.","Item first ignited, other","Item, Other","Liquids, piping, filters, other","Multiple items first ignited","Oily rags","Packing, wrapping material","Pipe, duct, conduit, hose covering","Pyrotechnics, explosives","Rubbish, trash, or waste","Rubbish, trash, waste","Storage supplies, other","Tire","Undetermined",NA))
```

### re-run missing data analysis
```{r}
mice::md.pattern(fires, rotate.names=TRUE)

# Keep only complete cases
fires <- fires[complete.cases(fires), ]
```

### splitting into train, tune, test
```{r}
set.seed(3001)
part_index_1 <- caret::createDataPartition(fires$IncidentDurationMinutes,
                                           times=1,
                                           p = 0.80,
                                           groups=1,
                                           list=FALSE)

train <- fires[part_index_1, ]
tune_and_test <- fires[-part_index_1, ]

#We need to use the function again to create the tuning set 
set.seed(3001)
tune_and_test_index <- createDataPartition(tune_and_test$IncidentDurationMinutes,
                                           p = .5,
                                           list = FALSE,
                                           times = 1)

tune <- tune_and_test[tune_and_test_index, ]
test <- tune_and_test[-tune_and_test_index, ]


dim(train)
dim(test) 
dim(tune)
```

## Decision Tree model

### Creating and training the model
```{r}
features <- train[, -5] # Remove incident duration since it's the target variable

target <- train$IncidentDurationMinutes

str(features)
str(target)

# We want to do 10-fold cross validation 5 times
fitControl <- trainControl(method = "repeatedcv",
                          number = 50,
                          repeats = 5)

# train the model
set.seed(1984)
duration_model_dt <- train(x=features,
                            y=target,
                            method="rpart2",
                            trControl=fitControl,
                            metric="RMSE")
```

### Running and Evaluating the model
```{r}
duration_model_dt

plot(duration_model_dt)

varImp(duration_model_dt)

rpart.plot(duration_model_dt$finalModel, type=5,extra=101)

grade_pred_reg <- predict(duration_model_dt, tune)

postResample(pred = grade_pred_reg, obs = tune$IncidentDurationMinutes)

range(tune$IncidentDurationMinutes)
555.3 - 6.4

# Get the nrmse
duration_dt_rmse <- postResample(pred = grade_pred_reg, obs = tune$IncidentDurationMinutes)[1]

(duration_dt_rmse / (555.3-6.4))
```

### Adjusting the hyperparameter
```{r}
tree.grid <- expand.grid(maxdepth=c(3:20))

# training the model with the adjusted hyperparameter
set.seed(1984)
duration_model_dt_1 <- train(x=features,
                            y=target,
                            method="rpart2",
                            trControl=fitControl,
                           tuneGrid=tree.grid,
                            metric="RMSE")
```

### Running and Evaluating the new model
```{r}
duration_model_dt_1

plot(duration_model_dt_1)

varImp(duration_model_dt_1)

rpart.plot(duration_model_dt_1$finalModel, type=5,extra=101)

grade_pred_reg <- predict(duration_model_dt_1, tune)

postResample(pred = grade_pred_reg, obs = tune$IncidentDurationMinutes)

range(tune$IncidentDurationMinutes)
555.3 - 6.4

# Get the nrmse
duration_dt_rmse_1 <- postResample(pred = grade_pred_reg, obs = tune$IncidentDurationMinutes)[1]

(duration_dt_rmse_1 / (555.3-6.4))
```

## Random Forest model - IGNORE EVERYTHING BELOW THIS

### Calculating the initial mtry level 
```{r}
mytry_tune <- function(x){
  xx <- dim(x)[2]-1
  sqrt(xx)
}


mytry_tune(fires)
# 3.87 = ~4

```

## Run the initial RF model with 1000 trees 
```{r}

fires_RF = randomForest(IncidentDurationMinutes~.,          #<- Formula: response variable ~ predictors.
                            #   The period means 'use all other variables in the data'.
                            train,     #<- A data frame with the variables to be used.
                            #y = NULL,           #<- A response vector. This is unnecessary because we're specifying a response formula.
                            #subset = NULL,      #<- This is unnecessary because we're using all the rows in the training data set.
                            #xtest = NULL,       #<- This is already defined in the formula by the ".".
                            #ytest = NULL,       #<- This is already defined in the formula by "PREGNANT".
                            ntree = 1000,        #<- Number of trees to grow. This should not be set to too small a number, to ensure that every input row gets classified at least a few times.
                            mtry = 5,            #<- Number of variables randomly sampled as candidates at each split. Default number for classification is sqrt(# of variables). Default number for regression is (# of variables / 3).
                            replace = TRUE,      #<- Should sampled data points be replaced.
                            #classwt = NULL,     #<- Priors of the classes. Use this if you want to specify what proportion of the data SHOULD be in each class. This is relevant if your sample data is not completely representative of the actual population 
                            #strata = NULL,      #<- Not necessary for our purpose here.
                            sampsize = 100,      #<- Size of sample to draw each time.
                            nodesize = 5,        #<- Minimum numbers of data points in terminal nodes.
                            #maxnodes = NULL,    #<- Limits the number of maximum splits. 
                            importance = TRUE,   #<- Should importance of predictors be assessed?
                            #localImp = FALSE,   #<- Should casewise importance measure be computed? (Setting this to TRUE will override importance.)
                            proximity = FALSE,    #<- Should a proximity measure between rows be calculated?
                            norm.votes = TRUE,   #<- If TRUE (default), the final result of votes are expressed as fractions. If FALSE, raw vote counts are returned (useful for combining results from different runs).
                            do.trace = TRUE,     #<- If set to TRUE, give a more verbose output as randomForest is run.
                            keep.forest = TRUE,  #<- If set to FALSE, the forest will not be retained in the output object. If xtest is given, defaults to FALSE.
                            keep.inbag = TRUE)   #<- Should an n by ntree matrix be returned that keeps track of which samples are in-bag in which trees? 

```

## Take a look at the variable importance measures, are they the same as the DT version or not? 
```{r}

fires_RF

fires_RF$call

fires_RF$rsq

workers_RF_acc = sum(workers_RF$confusion[row(workers_RF$confusion) == 
                    col(workers_RF$confusion)]) /sum(workers_RF$confusion)
# over 53% of people making >50k are misclassified

workers_RF_acc
# Accuracy was 0.84ish

View(as.data.frame(importance(fires_RF, type = 2, scale = TRUE)))
View(as.data.frame(fires_RF$importance))
# Marital status, relationship and occupation were the variables that had the largest mean decrease in accuracy.
# Occupation, Education and relationship led to the biggest mean decreases in Gini index.
# The most important variables for the decision tree were Capital Gain, Education and Occupation.
# It is interesting to note that Capital Gain, which was the most important variable in the decision tree by far, does not factor in the top 3 for either metric for the RF.
# Race, sex and native country were still among the least important variables in all cases.

# predict on the tune set
workers_predict_tune = predict(fires_RF,
                          tune,
                          type = "response",
                          predict.all = TRUE)

# confusion matrices of the two models
workers_RF$confusion
workers_RF_improved$confusion

library(caret)
confusionMatrix(as.factor(workers_predict_tune$aggregate),as.factor(tune$salary), 
                dnn=c("Prediction", "Actual"), mode = "everything")
# ~84% accuracy
# Sensitivity : 0.4761         
# Specificity : 0.9579

```